\texttt{dispRity} demo for ecologists

### 2016-03-15 doc/dispRity-ecology-demo.Rnw
Ecological demo vignette update

%1-Introduce the data
%"the data represents experimental plots in different treatments at two different water depths"

%2-Explain rarefaction better
%"the data represents experimental plots in different treatments at two different water depths"

%3-Hypothesis
%We want to know whether the treatments had an effect on the density of different species in the plots (Does treatment alter invertebrate community composition?). And does depth have an effect on community composition in the experimental plots? Annnnd: Do the treatments affect the communities differently at different depths?

%(Personally I hate them as you know!)  Just use them as a way of visually showing that the 'multivariate' communities in the different treatments/ depths are different or similar and I guess which components explain most of the variance. They really aren't that powerful which is why I love your approach. You sold it so well to me, I think your best selling points (and indeed perhaps actually including a sentence at the start) are that you don't lose all the dimensions of the data like you do in PCA. And that you can keep the important ones in (again like you sold to me, that, some axes may be far more influential than a pca with throw up).

@@

To keep it short, this package allows the use of all of the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a sub-set of dimensions.
For example, one might want to know whether the sort of water treatment at certain depth alters invertebrate communities and composition in natural habitats.

@@

This is a quick demo for using the \dispRity package (v.\Sexpr{packageVersion("dispRity")}) in ecological analysis.
See the \href{https://github.com/TGuillerme/dispRity}{other dispRity demos} for a general demo of the \dispRity package.

@@

## As well as two list of different factors affecting each experimental plot:
## the treatment and the depth.
treatments <- McClean_data$treatment
depth <- McClean_data$depth
@

@@

<<PCAplot, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## The x and y axis represent the two first dimensions of the ordination
x <- ord_matrix[, 1]
y <- ord_matrix[, 2]
## The colors will represent the treatments
cols <- sub("a", "red", treatments)
cols <- sub("b", "blue", cols)
## The symbols will represent the depth
pchs <- sub(1, 16, depth)
pchs <- as.numeric(sub(2, 17, pchs))

## Graphical option
par(bty = "n")
## A classic 2D ordination plot
plot(x, y, col = cols, pch = pchs, xlab = "Dimension 1", ylab = "Dimension 2",
    xlim = range(x) + c(0, 100))
@

This shows the distribution of the experimental plots along the two first axis of variance of the ordinated distance matrix (i.e. the first two dimensions).
At a first glance, it seems difficult to see a clear effect from the treatments or the depth since the different experimental plots with the same parameters don't seem to cluster.
However, the problem, is that this plot ignores the 18 other dimensions of the ordination!
Additionally, these two represented dimensions do not represent a biological reality \textit{per se}; i.e. the values on the first dimension do not represent some continuous traits (e.g. depth) but just the ordinations of correlations between the data and some factors.

Therefore, one might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a \textit{n}-dimensional object.
To explore our hypothesis, we could look to see if there are differences in the hyper-volume that different treatments occupy in the ordinated space to see the effect of water treatments on the communities.

@@

<<factors>>=
## Creating the table that contain the elements and their attributes
factors <- as.data.frame(matrix(data = c(treatments, depth), nrow = nrow(ord_matrix),
    ncol = 2, byrow = FALSE, dimnames = list(rownames(ord_matrix))))
names(factors)<-c("Treat", "Depth")
head(factors)
@

@@

## Displaying the full object
print.dispRity(customised_series, all=TRUE)
@

@@

In this example, we are going to define disparity as the median distance between the different experimental plots and the centroid of the ordinated space.
High values of disparity will indicate a general high dispersal from this centroid (i.e. on average, the experimental plots are far apart in the ordinated space).
We can define the metrics easily in the \dispRity function by feeding them to the \texttt{metric} argument.
Here we are going to feed the functions \texttt{median::stats} and \texttt{centroids::dispRity} which calculates distances between elements and centroid.

@@

<<boot.matrix>>=
## Simple bootstrapping (1000 times)
bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100)
## Bootstrapping with rarefaction (i.e. only re-sampling 17 elements each time)
rarefied_data <- boot.matrix(customised_series, bootstraps = 100, rarefaction = 17)
## Note that the output is a dispRity object giving some details on the series and the bootstraps
bootstrapped_data
rarefied_data
@

@@

<<dispRity3>>=
## Bootstrapped disparity
disparity_BS <- dispRity(bootstrapped_data, metric = c(median, centroids))
## Rarefied disparity
disparity_rare <- dispRity(rarefied_data, metric = c(median, centroids))
@

@@

<<summary>>=
## Summarising disparity (default)
summary(disparity_BS)
## The quantiles are calculated as 50 and 95 and the central tendency is the mean by default.
## But we can specify different options
summary(disparity_BS, quantile=90, cent.tend=median)

## Finally we can see the results of the rarefaction analysis:
summary(disparity_rare)
## Note that for the first three series, there is no observed disparity since the data was
## rarefied for these series.
@

@@

<<rare plots, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
## Graphical options
quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")
## Plotting the bootstrapped disparity
plot(disparity_BS, main="Bootstrapped data")
## Plotting the rarefied disparity
plot(disparity_rare, main="Rarefied data")
@

@@

<<test>>=
## Testing the effect of our factors on the bootstrapped data
summary(test.dispRity(disparity_BS, test = aov, comparisons = "all"))
## Testing the effect of our factors on the rarefied data
summary(test.dispRity(disparity_rare, test = aov, comparisons = "all"))

## Post-hoc testing of the effect of the two different factors
## Note that the comparison contains the list of the pairs of series to compare
test.dispRity(disparity_BS, test = wilcox.test, comparison = list(c(1,2), c(3,4)))
## Testing the effect of the two different factors for the rarefied data
test.dispRity(disparity_rare, test = wilcox.test, comparison = list(c(1,2), c(3,4)))
@

>>>>

@@

To keep it short, this package allows the use of all of the dimensions of ordinated matrices (i.e. PCA, MDS, PCO) for statistical analysis rather than just a sub-set of dimensions.
For example, one might want to know whether a specific kind of water treatment at certain water depth alters invertebrate communities and composition in natural habitats.

@@

@@

## As well as two list of different factors affecting each experimental plot:
## the treatment and the depth.
treatments <- McClean_data$treatment
depth <- McClean_data$depth

@

@@

<<PCAplot, fig.width=8, fig.height=8, out.width='.8\\linewidth'>>=
## The x and y axis represent the two first dimensions (i.e. principal
## components) of the ordination
x <- ord_matrix[, 1]
y <- ord_matrix[, 2]
## The colors will represent the treatments
cols <- sub("a", "red", treatments)
cols <- sub("b", "blue", cols)
## The symbols will represent the depth
pchs <- sub(1, 16, depth)
pchs <- as.numeric(sub(2, 17, pchs))

## Graphical option
par(bty = "n")
## A classic 2D ordination plot
plot(x, y, col = cols, pch = pchs, xlab = "PC 1", ylab = "PC 2",
    xlim = range(x) + c(0, 100))
@

This shows the distribution of the experimental plots along the two first axis of variance of the ordinated distance matrix (i.e. the first two dimensions).
At a first glance, it seems difficult to see a clear effect from the treatments (blue or red) or the depth (rounds or triangles) since the different experimental plots with the same parameters don't seem to cluster.
However, the problem, is that this plot ignores the 18 other dimensions of the ordination!
Additionally, these two represented dimensions do not represent a biological reality \textit{per se}; i.e. the values on the first dimension do not represent some continuous traits (e.g. depth) but just the ordinations of correlations between the data and some factors.

Therefore, one might want to approach this problem without getting stuck in only two dimensions and consider the whole dataset as a \textit{n}-dimensional object.

@@

<<factors>>=
## Creating the table that contain the elements and their attributes
factors <- as.data.frame(matrix(data = c(treatments, depth), nrow = nrow(ord_matrix),
    ncol = 2, byrow = FALSE, dimnames = list(rownames(ord_matrix))))
names(factors) <- c("Treat", "Depth")
head(factors)
@

@@

## Displaying the full object
print(customised_series, all = TRUE)
@

@@

In this example, we are going to define disparity as the median distance between the different experimental plots and the centroid of the ordinated space.
High values of disparity will indicate a general high dispersal from this centroid (i.e. on average, the experimental plots are far apart in the ordinated space).
We can define the metrics easily in the \dispRity function by feeding them to the \texttt{metric} argument.
Here we are going to feed the functions \texttt{stats::median} and \texttt{dispRity::centroids} which calculates distances between elements and centroid.

@@

<<boot.matrix>>=
## Simple bootstrapping (100 times)
bootstrapped_data <- boot.matrix(customised_series, bootstraps = 100)

## Bootstrapping with rarefaction (i.e. only re-sampling 17 elements each time)
rarefied_data <- boot.matrix(customised_series, bootstraps = 100,
    rarefaction = 17)

## Note that the output is a dispRity object giving some details on the series
## and the bootstraps
bootstrapped_data
rarefied_data
@

@@

<<dispRity3>>=
## Bootstrapped disparity
disparity_BS <- dispRity(bootstrapped_data, metric = c(median, centroids))

## Rarefied disparity
disparity_rare <- dispRity(rarefied_data, metric = c(median, centroids))
@

@@

<<summary>>=
## Summarising disparity (default)
summary(disparity_BS)

## The quantiles are calculated as 50 and 95 and the central tendency is the
## mean by default but we can specify different options
summary(disparity_BS, quantile = 90, cent.tend = median)

## Finally we can see the results of the rarefaction analysis:
summary(disparity_rare)

## Note that for the first three series, there is no observed disparity since
## the data was rarefied for these series.
@

@@

<<rare plots, fig.width=10, fig.height=5, out.width='1\\linewidth'>>=
## Graphical options
quartz(width = 10, height = 5) ; par(mfrow = (c(1,2)), bty = "n")

## Plotting the bootstrapped disparity
plot(disparity_BS, main = "Bootstrapped data")

## Plotting the rarefied disparity
plot(disparity_rare, main = "Rarefied data")
@

@@

<<test>>=
## Testing the effect of our factors on the bootstrapped data
summary(test.dispRity(disparity_BS, test = aov, comparisons = "all"))

## Testing the effect of our factors on the rarefied data
summary(test.dispRity(disparity_rare, test = aov, comparisons = "all"))

## Post-hoc testing of the effect of the two different factors
## Note that the comparison contains the list of the pairs of series to compare
test.dispRity(disparity_BS, test = wilcox.test,
    comparisons = list(c(1,2), c(3,4)))

## Testing the effect of the two different factors for the rarefied data
test.dispRity(disparity_rare, test = wilcox.test,
    comparisons = list(c(1,2), c(3,4)))
@
